Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-06-08 Manuscript/MPESubmission/response.tex

See our detailed response below.

\textcolor{blue}{3. Does the manuscript adhere to the PLOS Data Policy?}

\textcolor{blue}{Reviewer $\#$ 1: No}

We apologise for the confusion here. We mentioned where our data was stored in the supporting information but forgot to add this under the ``Data Reproducibility and Availability" section in the manuscript. We have now corrected this. All the data generated by the simulations are directly available on \href{http://figshare.com/articles/Effect_of_missing_data_on_topological_inference_using_a_total_evidence_approach/1306861}{FigShare} (goo.gl/h5Kxb5). Note that along with the data we share the full code used for repeating the submitted manuscript, from simulating the data to generating the figures and compiling the manuscript available on \href{https://github.com/TGuillerme/Total_Evidence_Method-Missing_data}{GitHub} (goo.gl/ewRsrm). All this code is provided with a detailed description of each step that we are happy to develop if needed.

@@

\textcolor{blue}{Reviewer $\#$ 1: This paper tackles an interesting questions. Despite some problems noted below, the experimental design seems adequate.}

%-------------

\textcolor{blue}{A - Unfortunately the data analysis performed is simply the calculation of a number of summary statistics about the overlap between the estimates returned for different simulation conditions. I suspect that if the authors take some time to do statistics (calculate some P-values, confidence intervals, or take a Bayesian perspective), they'll be able to make a statistical case for most of the points that they are trying to make. But I just don't think that a simulation comparison should be publishable without *some* statistical analysis (beyond the calculation of summaries).}

We feel that the reviewer has slightly misunderstood the statistics we applied in this manuscript. The purpose of ANOVAs or multiple regression (see suggestion from comment 20 below) is to compare differences in means across groups and test whether they are different from what one would expect according to a null model. In contrast, in our simulation approach, we essentially have access to a representative sample of the entire population distribution of tree topology comparisons. Therefore, we can simply ask whether these distributions are different when our missing data parameters varies by comparing them directly. One standard such approach to comparing distributions directly is to use the Bhattacharyya Coefficient. This number has direct interpretation in a probablistic sense and is simply the probability that the two distributions are the same. If the two distributions overlap completely then they have Bhattacharyya Coefficient and probability of being the same of 1. At the other extreme, if the two distributions are completely disjointed, then they have a Bhattacharyya Coefficient and hence probability of being the same of 0 (line 306 and 307 in the previously submitted manuscript and lines 320 and 319 in our revision). Thus, the Bhattacharyya Coefficient is a probability, comparable to, but not the same as, a p-value that could be calculated using frequentist approaches such as t-tests, ANOVAs or GLMs.

@@

Furthermore, as a direct comparison of two known distributions, using the the Bhattacharyya Coefficient comes with far fewer statistical assumptions than are required in ANOVAs, GLMs and their analogues. For example, the direct comparison approach using Bhattacharyya Coefficient is capable of comparing a multi-modal distributions whereas the assumption of normality required in a ANOVAs or GLMs thus using them would inviolate the robustness of the resultant p-value. There is an additional complication to using ANOVAs or GLMs in simulation type analyses arising from having access to very large sample sizes (e.g. 50000 data points for the Maximum Likelihood bootstraps trees and the Bayesian posterior distribution trees) that leads to p-values shrinking to zero and resulting in comparisons being almost certain to yield ``statistically significant" differences in the mean (see the figure below; code for reproducing it is available from \href{https://github.com/TGuillerme/Total_Evidence_Method-Missing_data/blob/master/Analysis/Comparing_distribution_example.R}{https://goo.gl/4djNUf}). We argue that in this situation, it is better to compare the entire distribution using the Bhattacharyya Coefficient rather than simply the means. Additionally, because we are comparing topologies in this manuscript, we made the methodological choice to use the modal topology (i.e. the most represented topology) instead of the mean topology (i.e. a hypothetical ``middle'' topology).

Finally, we would like to underline that we report our 50\% and 95\% confidence intervals in Figures 4 and 5 when showing the effect of the missing data parameters (calculation detailed lines 295-295) as well as the 1st and 3rd quartiles when comparing the methods in Table 1.

\includegraphics[scale=0.8, keepaspectratio=true]{Comparing_distributions.pdf} \\
\textit{The figure shows two extremely similar distributions. However, although the Bhattacharyya Coefficient shows that the distributions are really similar (0.999 overlap probability), a frequentist approach (simple t-test) suggests that there is a significant difference between the means (p-value $<$ 0.0000001). This highlights the reason we chose to use the Bhattacharyya Coefficient in our manuscript.}

@@

%-------------

\textcolor{blue}{B - The bias in the analysis in favor of Bayesian methods makes the recommendation to use Bayesian methods less satisfying.}

The reviewer is (we believe) referring to comment 13 below. We would like to emphasize that in this study we focused on the effect of missing data in the morphological part of the matrix on topology only, and we are ignoring branch lengths. The choice of the following priors (``exponential prior on the shape of the gamma distribution of $\alpha$ = 0.5 for both partitions" (line 228 in the previously submitted manuscript and line 240 in our revision) and the ``transition/transversion ratio prior of two sampled from a strong beta 229 distribution ($\beta$(80,40))" (line 229 in the previously submitted manuscript and line 241 in our revision)) affected mainly the molecular part of the matrix that remains constant (gamma distribution and transition/transversion ratio) and were a minor part of the MCMC sampler (2.17\% of the moves of the MCMC sampler for the three priors combined). Therefore we are confident that the topology is minimally affected by the priors, allowing us to explore the effect of missing morphological data on topology as well as reducing our computational time to a realistic 1.4 CPU centuries. Note that the use of the ``the ``true" tree's topology'' (line 226 in the previously submitted manuscript and line 237 in our revision) as a starting tree is not a prior and was used to avoid the chains getting stuck in local optima.

@@

Throughout our manuscript, we referred the M\textit{k} model \textit{sensu} Lewis 2001 (a Q-matrix with \textit{k} character states and an equal substitution rate $\mu$). The M\textit{kv} model is identical to the M\textit{k} model (as described in Lewis 2001) however, in practice, it assumes that all characters are variable and that therefore, one should correct for acquisition bias that can induce branch length inflation. This is generally done by using an algorithm in the phylogenetic software that adds ``dummy'' morphological characters. In some software (e.g. GARLI or BEAST), it is important to specify whether the morphological data should be corrected for acquisition bias or not (M\textit{k} or M\textit{kv} options). However, in RAxML and MrBayes, this bias is corrected automatically by addition of ``dummy'' characters and M\textit{k} just designates the evolutionary model used with $\mu$ $\geq$ 0 (e.g. Nylander \textit{et al.} 2004 - Systematic Biology).

The reviewer is of course correct in that real data rarely contain invariable morphological characters, because these don't contribute to phylogenetic signal. Our simulated data, on the other hand, will occasionally contain invariable characters merely because of missing data, therefore we made the decision to describe the model for morphological character evolution as the M\textit{k} evolutionary model ($\mu$ $\geq$ 0 rather than the M\textit{kv} model where $\mu$ $>$ 0). RAxML and MrBayes do however correct for acquisition bias as implemented in their algorithm for dealing with morphological data.

However, following the reviewer's comment, we changed the designation of the model from M\textit{k} to M\textit{kv} in the text, to underline that we corrected for acquisition bias. We also explicitly mentioned it in the revision in the following lines:

Lines 208 to 209 in our revision:

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
We used RAxML because it automatically corrects for acquisition bias (Lewis 2001). It is also heavily used in the literature for Maximum Likelihood tree inference (e.g. [...]) and is one of the fastest methods available (Stamatakis 2008).
\end{minipage}

Lines 233 to 234 in our revision:

@@

Both birth ($\lambda$) and death ($\mu$) parameters are probabilities bounded between 0 and 1. We have added this to the new version of the manuscript.

%-------------

\textcolor{blue}{5 - line 118: with 25 living and 25 fossil taxa. Are the fossils just put at the extinction point for the extinct lineages?}

Yes. We have clarified as follows in our revision (lines 118 to 119):

@@

See our response to comment C above.

%-------------

\textcolor{blue}{9 - line 171 - 177: the authors need to explain the order of operations for the $M_F$ vs $M_C$ masking. It makes a difference. If you mask out $M_F$ percentage of cells, and then mask $M_C$ percentage of columns, it is possible (indeed very probable) that the percentage of missing data in fossil characters that remain will not be $M_F$ after some of the morphological columns are removed. If you mask by $M_C$ first, and then $M_F$, the resulting matrices will always have $M_F$ percentage missing data (modulo rounding error) in the simulated realizations.}

We clarified this point ($M_C$ comes second) in the new version of the manuscript (lines 173 to 183). We also clarified the potential confusion between the amount of data removed in one parameter ($M_F$ = 10\% for example) and the amount of missing data in the morphological matrix (5\%) in the same revised paragraph (lines 173 to 183).

@@

Our simulations were designed to test the effects of missing data in as realistic a manner as possible given computational constraints. Thus, we do not feel that this additional analysis would add much to the manuscript, especially since it is already rather long and complex.

%-------------

\textcolor{blue}{12 - Filtering the data sets for strong bootstrap support under the full data analysis surely introduces some bias in favor of the no-missing data analyses.}

We implemented this step of keeping trees with a minimum total median bootstrap to make the topological comparisons more robust. If the no-missing data tree has low node support, it is likely that the changes in topology for the missing data trees are not linked to our focal question (the effect of removing characters, living taxa or fossils) but just to the low supported nodes. In fact, because we repeat the tree inference 125 times ($5^3$ parameters combinations), it is likely that we have 125 different pseudo-random solution for the nodes with really low phylogenetic support and that, therefore, the observed changes in topologies are just linked to the pseudo-random replication. Note however that we did not exclude this scenario from our protocol; the trees selected with a median bootstrap $>$ 50 could well contain some unresolved nodes with a low bootstrap ($<$ 50). We've added the following justification in our revision (lines 215 to 220):

@@

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
The Robinson-Foulds distance, or ``path difference", measures the difference between the number of clades and the twice number of shared clades across two trees.
\end{minipage}

%-------------

\textcolor{blue}{15 - line 258: RF is 0 (not 1) when the trees are identical. It's maximum (for 2 rooted binary trees) is 2(n-2) not n-2.}

We thanks the reviewer for pointing out this error and we changed the following sentence (lines 271 to 273 in our revision):

@@

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
This metric is bounded between zero, when the two trees are identical, and $2(n-2)$ (for two trees with $n$ taxa) when there is no shared clade in the two trees.
\end{minipage}

%-------------

\textcolor{blue}{16 - The ``scaling" to produce the Normalized RF ``distance" is confusing. It is also inverting the sign of the distance, so it has become a measure of similarity.}

We agree that this is confusing and thank the reviewer for pointing this out. We changed the mentions of ``Normalised Robinson-Foulds distance'' and ``Normalised Triplets distance'' to ``Normalised Robinson-Foulds metric'' and ``Normalised Triplets metric'' throughout the our revision. Additionally, we clarified the scaling method as follows (lines 276 to 284 in our revision):

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
We normalised this metric following Bogdanowicz \textit{et al.} (2012)'s Normalised Tree Similarity (NTS) method. This methods scales any tree comparison metric using the mean distance between 1000 random trees (see Supporting Information S2: Tree Comparisons for the calculation details). This method is a generalisation of the topological accuracy method (Price \textit{et al.} 2010) allowing to compare topological differences between any tree with any tree comparison metric. In practice when the Normalised Robinson-Foulds metric between two trees is equal to one, the trees are identical; if the metric is equal to zero, the trees are no more different than expected by chance; finally if the metric is less than zero, the trees are more different than expected by chance. Note that once rescaled, the Normalised Robinson-Foulds metric is a measure of similarity, rather than of distance like the original Robinson-Foulds metric. 
\end{minipage}
%-------------

@@

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
It is bounded between zero when the two trees are identical and $\binom{n}{3}$ (for two trees with $n$ taxa) when there is no shared taxa/clade position in the two trees.
\end{minipage}

%-------------

\textcolor{blue}{18 - Using the BC stat may be similar in spirit to using a t-test, but it is not equivalent.}

We clarified this point in the text as well as in the figures 2 and 3 captions (``equivalent" $->$ ``comparable" ; lines 324 to 326 in our revision):

@@

\textcolor{blue}{21 - line 398: ``mMthods" $->$ ``Methods"}

We thanks the reviewer for pointing out this typo and fixed it.

\textcolor{blue}{22 - line 420: contra the statement here, a consensus collapsing to a polytomy will affect the RF distance (could go up or down, but it will change).}

We revised this statement at lines 436 to 444 in our revision:

### 2015-06-08 Manuscript/MPESubmission/OnlineAppendices-LaTeXSuppFiles/SupportingInformation2.tex
Initial commit of MPE version

\subsection{Normalised Tree Similarity}
For any tree with \textit{n} taxa compared using a tree distance metric $m$, Normalized Tree Similarity, $NTS_m$ \cite{Bogdanowicz2012}, represents the similarity score for the two trees given the expected distance between two random Yule trees \cite{Bogdanowicz2012} with $n$ taxa. If $\bar{d}_{m,n}$\textit{(rand)} is the average distance between two random Yule trees with $n$ taxa and $d_{m,n}$\textit{(x,y)} the distance between the two trees \textit{x} and \textit{y} each containing $n$ taxa, then:
\begin{equation}
NTS_{m,n}(x,y)=\frac{\bar{d}_{m,n}(rand) - d_{m,n}(x,y)} {\bar{d}_{m,n}(rand)}
\end{equation}
\textit{NTS} ranges from one to -$\infty$.
For any $m,n$, when $NTS$ = 1, the trees are identical, when \textit{NTS} = 0 the trees are no more different than expected by chance, and when $NTS$ $<$ 0, the trees are more different than expected when comparing two random trees.

We used the NTS method to scale all the Robinson Foulds and Triplets distances calculated in our analyses, using the TreeCmp JavaScript \cite{Bogdanowicz2012}.

>>>>

\subsection{Normalised Tree Similarity}
For any tree with \textit{n} taxa compared using a tree difference metric $m$, Normalized Tree Similarity, $NTS_m$ \cite{Bogdanowicz2012}, represents the similarity score for the two trees given the expected difference between two random Yule trees \cite{Bogdanowicz2012} with $n$ taxa. If $\bar{d}_{m,n}$\textit{(rand)} is the average difference between two random Yule trees with $n$ taxa and $d_{m,n}$\textit{(x,y)} the difference between the two trees \textit{x} and \textit{y} each containing $n$ taxa, then:
\begin{equation}
NTS_{m,n}(x,y)=\frac{\bar{d}_{m,n}(rand) - d_{m,n}(x,y)} {\bar{d}_{m,n}(rand)}
\end{equation}
\textit{NTS} ranges from one to -$\infty$.
For any $m,n$, when $NTS$ = 1, the trees are identical, when \textit{NTS} = 0 the trees are no more different than expected by chance, and when $NTS$ $<$ 0, the trees are more different than expected when comparing two random trees.

We used the NTS method to scale all the Robinson Foulds and Triplets metrics calculated in our analyses, using the TreeCmp JavaScript \cite{Bogdanowicz2012}.
